Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2016-03-16 rmarkdown/paper.Rnw
added qc definiions to methods and also updated the section on chip-seq qc metrics applied to chip-exo

\subsection{Deeply sequence E. Coli $\sig$ ChIP-exo and ChIP-Seq data}

$\sig$ factor is a transcription initiation factor of housekeeping
genesin E. Coli. In this organism's genomes, many promoters contain
multiple transcription start sites (TSS) and these TSS are often
closely spaced (10 $\sim$ 150 bp). These closely spaced binding sites
are considered to be multiple ``switches'' that differentially
regulate gene expression under diverse growth conditions
\cite{regulondb}. Therefore, investigation of ChIP-exo's potential for
identification and differentiation of closely spaced binding sites are
invaluable for elucidating the transcriptional networks of prokaryotic
genomes.

\subsection{Current ChIP-Seq guidelines and quality metrics on ChIP-exo data}

We started our exploration by investigating the suitability of the
current state-of-the-art QC pipelines for ChIP-Seq for ChIP-exo. Table
\ref{tab:qc} contains measures that are commonly calculated for
ChIP-Seq samples: PCR Bottleneck Coefficient (PBC) and the Normalized
Strand Cross-Correlation (NSC) are calculated as in \cite{encode_qc}.
Additionally, we calculated the Forward Strand Ratio (FSR) to compare
the number of fragments sequenced from each strand.

@@

\begin{table}[h!]
%% need to fix format
  \centering
<<table1,include = FALSE,echo = FALSE,eval = TRUE>>=
  library(data.table)
  load("../data/paper_tables/Landick_rif_summary.RData")
  load("../data/paper_tables/Landick_aero_summary.RData")
  load("../data/paper_tables/pugh_ctcf.RData")
  load("../data/paper_tables/Carroll_mm9_FoxA1.RData")
  load("../data/paper_tables/Meijsing_hg19_GR.RData")
  load("../data/paper_tables/Carroll_hg19_ER.RData")
  landick_aero[,Organism := "E.Coli"]
  landick_aero[,Replicate := c(1,2,1,2)]
  landick_aero[,TF := "Sig70"]
  landick_aero[ ,cond := c(rep("Aerobic",2),rep("Anaerobic", 2))]
  landick_rif[,Organism := "E.Coli"]
  landick_rif[,Replicate := c(1,1,2,2)]
  landick_rif[,TF := "Sig70"]
  landick_rif[ ,cond := rep(c("Rif-0min","Rif-20min"),each = 2)]
  pugh_ctcf[,Organism := "Human"]
  pugh_ctcf[,Replicate := 1]
  pugh_ctcf[,TF := "CTCF"]
  pugh_ctcf[ ,cond := "HeLA"]
  carroll_mm9_FoxA1[,Organism := "Mouse"]
  carroll_mm9_FoxA1[,Replicate := c(3,1,2)]
  carroll_mm9_FoxA1[,TF := "Fox A1"]
  carroll_mm9_FoxA1[ ,cond := "Mouse Liver"]
  meijing_hg19_GR[,Organism := "Human"]
  meijing_hg19_GR[,Replicate := 1]
  meijing_hg19_GR[, TF := "GR"]
  meijing_hg19_GR[,cond := c("IMR90","K562","U20S")]
  carroll_hg19_ER[,Organism := "Human"]  
  carroll_hg19_ER[,Replicate := c(1,3,2)]
  carroll_hg19_ER[,TF := "ER"]
  carroll_hg19_ER[ ,cond := "MCF-7"]
  A <- do.call(rbind,list(landick_aero,
                     landick_rif[order(Replicate)],
                     pugh_ctcf[order(Replicate)],
                     carroll_mm9_FoxA1[order(Replicate)],
                     meijing_hg19_GR[order(Replicate)],
                     carroll_hg19_ER[order(Replicate)]))
  A[,nsc := unlist(nsc)]
  setcolorder(A, c("Organism","TF","cond","Replicate","nreads","pbc","nsc","files"))
  A[,files := NULL]
  setnames(A,names(A),c("Organism","IP/TF","Condition/Cell","Rep.","Depth","PBC","NSC"))
@ 
<<table2,include = TRUE,echo = FALSE, eval = TRUE >>= 
  A
@ 
\caption{Usual quality control indicators applied to the gathered
  ChIP-exo samples.  PBC stands for PCR Bottleneck Coefficient (0 -
  0.5 is severe bottlenecking, 0.5 - 0.8 is moderate bottlenecking,
  0.8 - 0.9 is mild bottlenecking, while 0.9 - 1 is no bottlenecking),
  NSC for Normalized Strand Cross-Correlation and FSR for Forward
  Strand Ratio. We omitted the Relative Strand Cross-Correlation (RSC)
  because a typical ChIP-exo experiment is not accompanied by an input
  file. $\sig$ samples were given by Dr. Landick's Lab, FoxA1 and ER
  samples are from Serandour et al., 2013 \cite{exoillumina} and CTCF
  sample is from Rhee and Pugh, 2011 \cite{exo1}}
  \label{tab:qc} %% need to make this table
\end{table}

It is of special importance to notice that the values of PBC are quite
low, hence it would be necessary to repeat these experiments by
following ChIP-seq guidelines. On the other hand, in a good ChIP-Seq
experiment the SCC is maximized at the fragment length, and a
``phantom peak'' would be also present at the experiment's read length
(\cite{encode_qc} and \cite{strandcc}). However, in ChIP-exo's case
these two peaks are confounded. Hence an enrichment measure based in
SCC such as the NSC is harder to interpret. In Figure
\ref{fig:scc_exo} we observe that both local maxima are hard to
differentiate in the SCC curves for the $\sig$ ChIP-exo datasets used to
calculate the NSC values in Table \ref{tab:qc}.

% % According to ENCODE's quality metrics, NSC values close
% % to one corresponds to low enrichment samples or bad quality datasets

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of \Sexpr{bin_size} non-overlapping
window after extending reads by \Sexpr{frag_len} to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

@@

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .9\textwidth]{../figs/for_paper/sig70_methods_comparison_FDR5_topM300.pdf}
  \caption{Comparison of the resolution between dPeak, Gem, Mace and
    Peakzilla methods. Resolution is defined as the minimum distance
    between a RegulonDB annotation an a predicted binding event.}
  \label{fig:methods_comp}
\end{figure}

>>>>

\subsection{Deeply sequenced E. Coli $\sig$ ChIP-exo and ChIP-Seq data}

$\sig$ factor is a transcription initiation factor of housekeeping
genes in E. Coli. In this organism's genomes, many promoters contain
multiple transcription start sites (TSS) and these TSS are often
closely spaced (10 $\sim$ 150 bp). These closely spaced binding sites
are considered to be multiple ``switches'' that differentially
regulate gene expression under diverse growth conditions
\cite{regulondb}. Therefore, investigation of ChIP-exo's potential for
identification and differentiation of closely spaced binding sites is
invaluable for elucidating the transcriptional networks of prokaryotic
genomes.

\subsection{Application of current ChIP-Seq QC guidelines on ChIP-exo
  data}

We started our exploration by investigating whether the current
state-of-the-art QC pipelines for ChIP-Seq are suitable for
ChIP-exo. Table \ref{tab:qc} contains measures that are commonly
calculated for ChIP-Seq samples: PCR Bottleneck Coefficient (PBC) and
the Normalized Strand Cross-Correlation (NSC) are calculated as in
\cite{encode_qc}. We omitted the Relative Strand Cross-Correlation
(RSC) which is another commonly used QC measure because a typical
ChIP-exo experiment is not accompanied by an input sample.

@@

\begin{table}[h!]
%% need to fix format
  \centering
<<table1,include = FALSE,echo = FALSE,eval = TRUE>>=
  library(data.table)
  load("../data/paper_tables/Landick_rif_summary.RData")
  load("../data/paper_tables/Landick_aero_summary.RData")
  load("../data/paper_tables/pugh_ctcf.RData")
  load("../data/paper_tables/Carroll_mm9_FoxA1.RData")
  load("../data/paper_tables/Meijsing_hg19_GR.RData")
  load("../data/paper_tables/Carroll_hg19_ER.RData")
  landick_aero[,Organism := "E.Coli"]
  landick_aero[,Replicate := c(1,2,1,2)]
  landick_aero[,TF := "Sig70"]
  landick_aero[ ,cond := c(rep("Aerobic",2),rep("Anaerobic", 2))]
  landick_rif[,Organism := "E.Coli"]
  landick_rif[,Replicate := c(1,1,2,2)]
  landick_rif[,TF := "Sig70"]
  landick_rif[ ,cond := rep(c("Rif-0min","Rif-20min"),each = 2)]
  pugh_ctcf[,Organism := "Human"]
  pugh_ctcf[,Replicate := 1]
  pugh_ctcf[,TF := "CTCF"]
  pugh_ctcf[ ,cond := "HeLA"]
  carroll_mm9_FoxA1[,Organism := "Mouse"]
  carroll_mm9_FoxA1[,Replicate := c(3,1,2)]
  carroll_mm9_FoxA1[,TF := "Fox A1"]
  carroll_mm9_FoxA1[ ,cond := "Mouse Liver"]
  meijing_hg19_GR[,Organism := "Human"]
  meijing_hg19_GR[,Replicate := 1]
  meijing_hg19_GR[, TF := "GR"]
  meijing_hg19_GR[,cond := c("IMR90","K562","U20S")]
  carroll_hg19_ER[,Organism := "Human"]  
  carroll_hg19_ER[,Replicate := c(1,3,2)]
  carroll_hg19_ER[,TF := "ER"]
  carroll_hg19_ER[ ,cond := "MCF-7"]
  A <- do.call(rbind,list(landick_aero,
                     landick_rif[order(Replicate)],
                     pugh_ctcf[order(Replicate)],
                     carroll_mm9_FoxA1[order(Replicate)],
                     meijing_hg19_GR[order(Replicate)],
                     carroll_hg19_ER[order(Replicate)]))
  A[,nsc := unlist(nsc)]
  setcolorder(A, c("Organism","TF","cond","Replicate","nreads","pbc","nsc","files"))
  A[,files := NULL]
  A[,nreads := prettyNum(nreads,big.mark = ",")]
  A[,pbc := round(pbc,4)]
  A[,nsc := round(nsc,4)]
  setnames(A,names(A),c("Organism","IP/TF","Condition/Cell","Rep.","Depth","PBC","NSC"))
@

<<table2,include = TRUE,echo = FALSE, eval = TRUE >>= 
  A
@ 
\caption{\textbf{Current QC metrics applied to ChIP-exo data.} PBC
  stands for PCR Bottleneck Coefficient and NSC for Normalized Strand
  Cross-Correlation.}
  \label{tab:qc} %% need to make this table
\end{table}

DNA libraries generated by the ChIP-exo protocol seem to be less
complex than the libraries generated by ChIP-Seq, since the possible
number of positions to which the reads can be aligned is being reduced
due to the exonuclease digestion, hence considerable amounts of reads
are being mapped to specific positions. This affects the
interpretation of the PBC, since for ChIP-Seq low PBC values indicate
that the same read has been copied by the amplification process and
aligned multiple times to the same position; while for ChIP-exo when
several reads are aligned to the same position are not necessarily the
same read amplified, but several reads that their $5\prime$ end was
digested to the same position before the amplification step. It is of
special importance to notice that for several ChIP-exo datasets, the
PBC values are quite low. Therefore, by blindly following ChIP-Seq
guidelines those experiment could have been considered as not useful
and repeated.

The Strand Cross-Correlation (SCC) introduced by Kharchenko et al.,
2008 \cite{strandcc} is the most commonly used quality measure in
ChIP-Seq. It is calculated as the correlation between both strand
coverages, where each one is shifted $\delta / 2$ bp towards the
$3\prime$ direction. In general it measures how well the reads mapped
to each strand are clustered around the locations where the proteins
are binding to the DNA, and usually it is expected to observed two
local maxima, one when the profiles are shifted by the average read
length and another when the profiles are shifted by the unobserved
fragment length. In a good ChIP-Seq dataset the last one is also the
SCC global maxima. However, in ChIP-exo's case these two peaks are
confounded. Hence the Normalized Strand Cross-Correlation (NSC) which
is a measure based on the SCC is harder to interpret. Figure
\ref{fig:scc_exo} shows that both local maxima are hard to
differentiate in the SCC curves for the $\sig$ ChIP-exo datasets used
to calculate the NSC values in Table \ref{tab:qc}.

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of \Sexpr{bin_size} non-overlapping
window after extending reads by \Sexpr{frag_len} to their $3\prime$
end directions. ChIP tag counts in ChIP-exo data were linearly related
to ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

@@

\subsection*{Definition of current ChIP-Seq QC guidelines.}

\subsubsection*{PCR Bottleneck Coefficient.}

The PCR Bottleneck coefficient is a measure of library complexity in
ChIP-Seq data:

\begin{align}
  \mbox{PBC} = \frac{\text{Nr. of positions to which exactly one
      unique mapping read is aligned}}{\text{Nr. of positions to
      which at least one unique mapping read is aligned}} \nonumber
\end{align}

For human and mouse genome, the ENCODE project states that a PBC value
in the 0 - 0.5 range indicates severe bottlenecking, in the 0.5 - 0.8
range moderate bottlenecking, in the 0.8 - 0.9 range indicates mild
bottlenecking and in the 0.9 - 1 range indicates that there is no
presence of bottlenecking.

\subsubsection*{Strand Cross-Correlation.}

The strand cross-correlation was proposed by Kharchenko et al., 2008
\cite{strandcc} and it may be one of the most used of the ChIP-Seq QC
metrics. The SCC curve is defined as:

\begin{align}
  y(\delta) = \sum_c w_c r\left[ n_c^+ \left(x + \frac{\delta}{2}
    \right), n_c^- \left( x- \frac{\delta}{2} \right)\right],
\label{scc}
\end{align}

where $y(\delta)$ is the SCC for a strand shift $\delta$, $r$ is the
Pearson correlation, $w_c$ is the proportion of reads mapped to
chromosome $c$ and $n_c^S$ is the read count vector for strand $S$ and
chromosome $c$. Additionally, two QC metrics are defined:

\begin{align}
  \mbox{NSC} &= \frac{\max_\delta y(\delta)}{\min_\delta y(\delta)}, \label{nsc} \\
  \mbox{RSC} &= \frac{\max_\delta y(\delta) -
    y_{\text{bgd}}}{y_{\text{rl}} - y_{\text{bgd}}}. \label{rsc}
\end{align}

where $y_{\text{bgd}}$ is the background SCC level and $y_{\text{rl}}$
is the SCC ``phantom peak'' value.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .9\textwidth]{../figs/for_paper/sig70_methods_comparison_FDR5_topM300.pdf}
  \caption{Comparison of the resolution between dPeak, Gem, Mace and
    Peakzilla methods. Resolution is defined as the minimum distance
    between a RegulonDB annotation and a predicted binding event.}
  \label{fig:methods_comp}
\end{figure}

### 2016-03-16 rmarkdown/paper.tex
added qc definiions to methods and also updated the section on chip-seq qc metrics applied to chip-exo

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
##     Organism  IP/TF Condition/Cell Rep.     Depth        PBC        NSC
##  1:   E.Coli  Sig70        Aerobic    1  13961493 0.13994203 103.423876
##  2:   E.Coli  Sig70        Aerobic    2  14810838 0.16339252 162.800188
##  3:   E.Coli  Sig70      Anaerobic    1  16108774 0.13535422 153.508809
##  4:   E.Coli  Sig70      Anaerobic    2  13636541 0.15322537 172.581549
##  5:   E.Coli  Sig70       Rif-0min    1    902921 0.26895580  13.769077
##  6:   E.Coli  Sig70       Rif-0min    1   1852124 0.25908459  17.918825
##  7:   E.Coli  Sig70      Rif-20min    2   2104427 0.25841987  29.608305
##  8:   E.Coli  Sig70      Rif-20min    2  11548572 0.15107930  13.086255
##  9:    Human   CTCF           HeLA    1  48478450 0.45796889  16.024817
## 10:    Mouse Fox A1    Mouse Liver    1  22210461 0.65622704  21.281972
## 11:    Mouse Fox A1    Mouse Liver    2  23307557 0.79962169  60.421912
## 12:    Mouse Fox A1    Mouse Liver    3  22421729 0.10683407  72.042428
## 13:    Human     GR          IMR90    1  47443803 0.29785605   8.860176
## 14:    Human     GR           K562    1 116518000 0.05047634   4.117883
## 15:    Human     GR           U20S    1   3255111 0.77142141  10.058834
## 16:    Human     ER          MCF-7    1   9289835 0.80825680  19.875232
## 17:    Human     ER          MCF-7    2  11041833 0.80241200  21.485094
## 18:    Human     ER          MCF-7    3  12464836 0.82035947  18.728062
\end{verbatim}
\end{kframe}
\end{knitrout}
\caption{Usual quality control indicators applied to the gathered
  ChIP-exo samples.  PBC stands for PCR Bottleneck Coefficient (0 -
  0.5 is severe bottlenecking, 0.5 - 0.8 is moderate bottlenecking,
  0.8 - 0.9 is mild bottlenecking, while 0.9 - 1 is no bottlenecking),
  NSC for Normalized Strand Cross-Correlation and FSR for Forward
  Strand Ratio. We omitted the Relative Strand Cross-Correlation (RSC)
  because a typical ChIP-exo experiment is not accompanied by an input
  file. $\sig$ samples were given by Dr. Landick's Lab, FoxA1 and ER
  samples are from Serandour et al., 2013 \cite{exoillumina} and CTCF
  sample is from Rhee and Pugh, 2011 \cite{exo1}}
  \label{tab:qc} %% need to make this table
\end{table}

It is of special importance to notice that the values of PBC are quite
low, hence it would be necessary to repeat these experiments by
following ChIP-seq guidelines. On the other hand, in a good ChIP-Seq
experiment the SCC is maximized at the fragment length, and a
``phantom peak'' would be also present at the experiment's read length
(\cite{encode_qc} and \cite{strandcc}). However, in ChIP-exo's case
these two peaks are confounded. Hence an enrichment measure based in
SCC such as the NSC is harder to interpret. In Figure
\ref{fig:scc_exo} we observe that both local maxima are hard to
differentiate in the SCC curves for the $\sig$ ChIP-exo datasets used to
calculate the NSC values in Table \ref{tab:qc}.

% % According to ENCODE's quality metrics, NSC values close
% % to one corresponds to low enrichment samples or bad quality datasets

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

>>>>

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
##     Organism  IP/TF Condition/Cell Rep.       Depth    PBC      NSC
##  1:   E.Coli  Sig70        Aerobic    1  13,961,493 0.1399 103.4239
##  2:   E.Coli  Sig70        Aerobic    2  14,810,838 0.1634 162.8002
##  3:   E.Coli  Sig70      Anaerobic    1  16,108,774 0.1354 153.5088
##  4:   E.Coli  Sig70      Anaerobic    2  13,636,541 0.1532 172.5815
##  5:   E.Coli  Sig70       Rif-0min    1     902,921 0.2690  13.7691
##  6:   E.Coli  Sig70       Rif-0min    1   1,852,124 0.2591  17.9188
##  7:   E.Coli  Sig70      Rif-20min    2   2,104,427 0.2584  29.6083
##  8:   E.Coli  Sig70      Rif-20min    2  11,548,572 0.1511  13.0863
##  9:    Human   CTCF           HeLA    1  48,478,450 0.4580  16.0248
## 10:    Mouse Fox A1    Mouse Liver    1  22,210,461 0.6562  21.2820
## 11:    Mouse Fox A1    Mouse Liver    2  23,307,557 0.7996  60.4219
## 12:    Mouse Fox A1    Mouse Liver    3  22,421,729 0.1068  72.0424
## 13:    Human     GR          IMR90    1  47,443,803 0.2979   8.8602
## 14:    Human     GR           K562    1 116,518,000 0.0505   4.1179
## 15:    Human     GR           U20S    1   3,255,111 0.7714  10.0588
## 16:    Human     ER          MCF-7    1   9,289,835 0.8083  19.8752
## 17:    Human     ER          MCF-7    2  11,041,833 0.8024  21.4851
## 18:    Human     ER          MCF-7    3  12,464,836 0.8204  18.7281
\end{verbatim}
\end{kframe}
\end{knitrout}
\caption{\textbf{Current QC metrics applied to ChIP-exo data.} PBC
  stands for PCR Bottleneck Coefficient and NSC for Normalized Strand
  Cross-Correlation.}
  \label{tab:qc} %% need to make this table
\end{table}

DNA libraries generated by the ChIP-exo protocol seem to be less
complex than the libraries generated by ChIP-Seq, since the possible
number of positions to which the reads can be aligned is being reduced
due to the exonuclease digestion, hence considerable amounts of reads
are being mapped to specific positions. This affects the
interpretation of the PBC, since for ChIP-Seq low PBC values indicate
that the same read has been copied by the amplification process and
aligned multiple times to the same position; while for ChIP-exo when
several reads are aligned to the same position are not necessarily the
same read amplified, but several reads that their $5\prime$ end was
digested to the same position before the amplification step. It is of
special importance to notice that for several ChIP-exo datasets, the
PBC values are quite low. Therefore, by blindly following ChIP-Seq
guidelines those experiment could have been considered as not useful
and repeated.

The Strand Cross-Correlation (SCC) introduced by Kharchenko et al.,
2008 \cite{strandcc} is the most commonly used quality measure in
ChIP-Seq. It is calculated as the correlation between both strand
coverages, where each one is shifted $\delta / 2$ bp towards the
$3\prime$ direction. In general it measures how well the reads mapped
to each strand are clustered around the locations where the proteins
are binding to the DNA, and usually it is expected to observed two
local maxima, one when the profiles are shifted by the average read
length and another when the profiles are shifted by the unobserved
fragment length. In a good ChIP-Seq dataset the last one is also the
SCC global maxima. However, in ChIP-exo's case these two peaks are
confounded. Hence the Normalized Strand Cross-Correlation (NSC) which
is a measure based on the SCC is harder to interpret. Figure
\ref{fig:scc_exo} shows that both local maxima are hard to
differentiate in the SCC curves for the $\sig$ ChIP-exo datasets used
to calculate the NSC values in Table \ref{tab:qc}.

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their $3\prime$
end directions. ChIP tag counts in ChIP-exo data were linearly related
to ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.
